E-Graphs With Bindings

### 2025-01-24 LICS_2025/e_graphs_with_bindings/sections/combinatorial_semantics.tex
sections done

First,  we fix some notation.  Let $(-)^*$ be the free monoid monad over \textsf{Set}.  
We extend this to the case of relations $R \subseteq V \times V$,  so that we denote by $R^{*} \subseteq V^* \times V^*$ the element-wise extension of $R$ to a relation on ordered sequences.
We will sometimes use functional notation for relations,  so that given $R \subseteq V \times W$ and $v \in V$,  $R(v) \subseteq W$. 
We elide associativity and unit isomorphisms associated with coproducts $(+)$,  and denote by $\iota_j: X_{j} \rightarrow X_{1} + \ldots X_{j} + \ldots X_{n}$ the $j^{\text{th}}$ injection,  and $[f,g]: A_1 + A_2 \to B$ the co-pairing of two morphisms $f: A_1 \to B$ and $g:A_2 \to B$.
We denote by $A +_{f,g} B$ the pushout of the span $A \xleftarrow{f} C \xrightarrow{g} B$.
We will also refer to $A,B$ as \textit{feet} of the cospan, and to $C$ as \textit{carrier}.
Given an ordered sequence of labels from some alphabet $\Sigma$, $[l_1, \ldots, l_n]$ we will denote with $w([l_1, \ldots, l_n])$ the word concatenated from these labels using $\otimes$, \textit{e.g.}, $w([A,B]) = A \otimes B$.

@@

In ~\cite{bonchi_string_2022-1},  hypergraphs \textit{with interfaces} and their DPO rewriting are shown to be sound and complete for categorical presentations of SMTs including a Frobenius algebra. 
Hypergraphs generalise graphs by allowing edges to have multiple sources and targets. 
When interpreting string diagrams as hypergraphs,  morphisms are interpreted as edges,  and wires as vertices.  
However,  there is also a need to specify which vertices are to be considered as \textit{input} and \textit{outputs},  corresponding to the dangling wires at the bottom and top of a string diagram. 
This gives rise to the definition of a hypergraph with interfaces.  
Subsequent work \cite{bonchi_string_2022-2} builds on this correspondence to give a restriction of hypergraphs with interfaces (called \textit{monogamous} directed acyclic),  leading to a correspondence between symmetric monoidal string diagrams (\textit{i.e.}, without requiring a Frobenius algebra) and the category of (cospans of) such hypergraphs.
We then evoke the definition of hypergraphs and corresponding cospans as introduced in~\cite{bonchi_string_2022-1,bonchi_string_2022-2} over not uni-typed signatures.

\begin{definition}[Category of hypergraphs]\label{def:hypergraph}
A \emph{hypergraph $\mathcal{G}$ over monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$} is a tuple $(V,E,s,t,l)$,  where $V$ is a finite set of vertices, $E$ is a finite set of edges, $s : E \to V^{*}$ is a source function, $t : E \to V^{*}$ is a target function,  and $l = (l_{V} : V \to \Sigma_{O},l_{E} : E \to \Sigma_{M})$ is a pair of labelling functions that assigns each edge and vertex an operator and a type from monoidal signature $\Sigma$ respectively.  
The labelling function must respect the typing of the operator: for all edges $e$, such that $l(e) = c : A \to B$,  $l_{V}^{*}(s(e)) = A$ and $l_{V}^{*}(t(e)) = B$.  
We call a hypergraph \textit{discrete} if its set of edges is empty.   
A \emph{hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ is given by a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that the following hold. 
\begin{enumerate}
    \item $\phi_V^*(s_{\mathcal{F}}(e)) = s_{\mathcal{G}}(\phi_E(e))$
    \item $\phi_V^*(t_{\mathcal{F}}(e)) = t_{\mathcal{G}}(\phi_E(e))$
    \item $l_{V,\mathcal{F}}(v) = l_{V,\mathcal{G}}(\phi_V(v))$
    \item $l_{E,\mathcal{F}}(e) = l_{E,\mathcal{G}}(\phi_E(e))$
\end{enumerate}
That is, a homomorphism should preserve sources and targets of edges and preserve the labels.
We denote by $\catname{Hyp(\Sigma)}$ the category of hypergraphs over $\Sigma$ and hypergraph homomorphisms.
\end{definition}
$\catname{Hyp(\Sigma)}$ has all finite colimits and, in particular,  the coproduct is given by the disjoint union of hypergraphs.
The initial object is given by the empty hypergraph.

@@

% Then, morphisms of $\catname{S}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces~\cite{bonchi_string_2022-2}.
\begin{definition}[Hypergraphs with interfaces]
\label{def:cspd}
A hypergraph with interfaces is a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\catname{Hyp}(\Sigma)$ where $n$, $m$ are discrete ordered hypergraphs that denote interfaces.
Define the category of \emph{hypergraphs with interfaces} $\catname{HypI(\Sigma)}$ as the category with discrete \textit{ordered} hypergraphs as objects and isomorphism classes of hypergraphs with interfaces as morphisms.
\end{definition}
$\HypI{\Sigma}$ is a symmetric monoidal category with the tensor product given by coproduct in $\catname{Hyp}(\Sigma)$:

\begin{align*}
n \xrightarrow{f} &\mathcal{F} \xleftarrow{g} m\\
&\times \qquad \quad \Coloneqq n + n' \xrightarrow{f + f'} \mathcal{F} + \mathcal{G} \xleftarrow{g + g'} m + m'\\
n' \xrightarrow{f'} &\mathcal{G} \xleftarrow{g'} m'
\end{align*}
Unit is given by the initial object of $\catname{Hyp}(\Sigma)$.
Symmetry is inherited from coproduct in $\catname{Hyp}(\Sigma)$~\cite{MonoidalCoproduct}.

Composition of two cospans $n \xrightarrow{f} \mathcal{F} \xleftarrow{g} m;m \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} k$ is given by the pushout
\[
n \xrightarrow{f} \mathcal{F} +_{g,f'} \mathcal{G} \xleftarrow{g'} k
\]

that is computed by identifying vertices in $\mathcal{F} + \mathcal{G}$ that share pre-image in $m$.
As pushout is only defined up to isomorphism, it justifies the morphisms in this category being isomorphism classes of cospans.

To make such cospans complete with respect to representing morphism of $\catname{S}(\Sigma)$, we need to impose some restriction in their form.

\begin{definition}[Degree,  path,  acyclic] 
The \emph{in-degree} of a vertex $v$ in a hypergraph $\mathcal{G}$ is the number of hyperedges $e \in {E_\mathcal{G}}$ such that $v \in t(e)$.  
Similarly, the out-degree of $v$ is the number of edges $e \in E_\mathcal{G}$ such that $v \in s(e)$.
A \emph{path} from $e_0$ to $e_{n-1}$ of length $n$ in a hypergraph $\mathcal{G}$ is an ordered sequence of edges $[e_0, e_1, \ldots, e_{n-1}] \in E_\mathcal{G}^*$ such that for all $i < n - 1$ some vertex $v \in t(e_i)$ is also in $s(e_{i+1})$.
A hypergraph is \emph{directed acyclic} if it contains no paths of length $n > 0$ that start and end in the same edge.
\end{definition}

@@

\begin{figure}[t!]
    \begin{subfigure}[T]{0.4\linewidth}
    \[
                % \tikzfig{../figures/combinatorial_semantics/DPO_square}
                \begin{tikzcd}
                    {\mathcal{L}} & {\mathcal{K}} & {\mathcal{R}} \\
                    {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}}
                    \arrow["f"', from=1-1, to=2-1]
                    \arrow[from=1-2, to=1-1]
                    \arrow[from=1-2, to=1-3]
                    \arrow[from=1-2, to=2-2]
                    \arrow[from=1-3, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
                    \arrow[from=2-2, to=2-1]
                    \arrow[from=2-2, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
                \end{tikzcd}
    \]
    % \subcaption{DPO square}
    % \label{fig:dpo}
    \end{subfigure}
    \hfill
    \begin{subfigure}[T]{0.4\linewidth}
        \[
    % \scalebox{0.8}{
        % \tikzfig{../figures/combinatorial_semantics/DPOI_square_Hyp}
    % }
    \begin{tikzcd}
        {\mathcal{L}} & {i+j} & {\mathcal{R}} \\
        {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}} \\
        & {n+m}
        \arrow["f"', from=1-1, to=2-1]
        \arrow[from=1-2, to=1-1]
        \arrow[from=1-2, to=1-3]
        \arrow[from=1-2, to=2-2]
        \arrow[from=1-3, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
        \arrow[from=2-2, to=2-1]
        \arrow[from=2-2, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
        \arrow[from=3-2, to=2-1]
        \arrow[from=3-2, to=2-2]
        \arrow[from=3-2, to=2-3]
    \end{tikzcd}
\]
    % \subcaption{DPOI square}
    % \label{fig:dpoi}
    \end{subfigure}
    \captionsetup{belowskip=-3ex}
    \caption{DPO and DPOI squares}
    \label{fig:dpo_dpoi}
\end{figure}

@@

Then, the authors of~\cite{bonchi_string_2022-2} showed that this enhanced notion of DPOI rewriting is sound and complete in the following sense.
Any two $\Sigma$-terms $f$ and $g$ are equal modulo SMC laws and equations $\mathcal{E}$ if and only if $\llbracket f \rrbracket \Rrightarrow{}_{\llbracket \mathcal{E} \rrbracket} \llbracket g \rrbracket$ where $\llbracket - \rrbracket$ is the functor giving the equivalence in Theorem~\ref{thm:prop-equiv} and $\llbracket E \rrbracket = \{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle, \langle \llbracket r \rrbracket, \llbracket l \rrbracket \rangle \text{ for } l=r \in \mathcal{E} \}$.
We then proceed with building a combinatorial representation for closed $\Sigma^{+}$-terms to obtain a notion of DPOI rewriting that would also subsume equality modulo SMC laws.

\section{Closed E-hypergraphs}
Prior work of~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} introduced a notion of \textit{e-hypergraph} --- a combinatorial representation of morphism terms for free semilattice-enriched symmetric monoidal category with a modified notion of DPOI rewriting to make it sound and complete for equalities between morphism terms.
The semilattice enrichment structure was encoded using hierarchical edges in a hypergraph following the approach of a well-established area of hierarchical hypergraphs~\cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical, fscd}.
The latter work in \textit{loc. cit.}, in particular, introduced a combinatorial representation of closed $\Sigma$-terms (without enrichment).
Notably, the laws of a semilattice enrichment in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and the laws of Definition~\ref{def:closed} in~\cite{fscd} were not absorbed by the combinatorial representation and instead were implemented on top as structural DPOI rewrites.

@@

Consider an example in the middle part of Figure~\ref{fig:e-cospan-example}.
The e-hypergraph in the example is given by $V = \{v_{1}, \ldots w_{10}\}$, $E = \{e_{1}, \ldots, e_{7}\}$, $\textcolor{red}{E} = \{e_{1}\}$, $\textcolor{blue}{E} = \{e_{2}\}$, $l_{E} = {e_{1} \mapsto \bot, e_{2} \mapsto \bot, e_{3} \mapsto g, \ldots, e_{7} \mapsto f}$ and then we have
\[
\begin{array}{cccc}
  v_{4} \textcolor{red}{<^{\mu}} e_{1} & v_{7} \textcolor{blue}{<^{\mu}} e_{2} & v_{4} \consistency e_{2} & v_{4} \not \consistency v_{10}\\
  e_{3} \textcolor{red}{<^{\mu}} e_{1} & v_{8} \textcolor{blue}{<^{\mu}} e_{2} &  e_{2} \consistency u_{1} & e_{2} \not \consistency e_{6}\\
  e_{2} \textcolor{red}{<^{\mu}} e_{1} & e_{5} \textcolor{blue}{<^{\mu}} e_{2} & e_{2} \consistency e_{4} & e_{4} \not \consistency e_{7}\\  
  \ldots & \ldots & \ldots & \ldots \\
  v_{12} \textcolor{red}{<^{\mu}} e_{1} & w_{9} \textcolor{blue}{<^{\mu}} e_{2} & v_{11} \consistency v_{12} & w_{6} \not \consistency w_{10}
\end{array}
\]
and, in particular, we have $[e_{5}) = \{e_2, e_1\}$.
$\consistency_{e_{1}}$ defines a partition on immediate successors of $e_{1}$ which we delimit by a vertical dashed line.
Edges in $\textcolor{red}{E}$ are depicted with a dashed box, while edges in $\textcolor{blue}{E}$ --- with a round solid box.

\begin{figure}

\[
\adjustbox{scale=0.6}{
\tikzfig{./figures/closed_iso_example_2}
}
\]
\caption{Cospan of e-hypergraphs example}
\label{fig:e-cospan-example}
\end{figure}

\begin{definition}[E-hypergraph homomorphism]
  An \textit{e-hypergraph homomorphism} $\phi:\mathcal{F}\to\mathcal{G}$ between e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
  \begin{enumerate}
      \item $\phi$ is a hypergraph homomorphism;
      \item $\phi_{E}(\textcolor{red}{E_{\mathcal{F}}}) \subseteq \textcolor{red}{E_{\mathcal{G}}}$ and $\phi_{E}(\textcolor{blue}{E_{\mathcal{F}}}) \subseteq \textcolor{blue}{E_{\mathcal{G}}}$;
      \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
        \begin{align*}
          &\text{if } e \textcolor{red}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{red}{<_{\mathcal{G}}^{\mu}} \phi(v)\\
          &\text{and}\\
          &\text{if } e \textcolor{blue}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{blue}{<_{\mathcal{G}}^{\mu}} \phi(v)
        \end{align*}
          and for $e_1, e_2 \in E_{\mathcal{F}}$
          \begin{align*}
          &\text{if } e_1 \textcolor{red}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{red}{<_{\mathcal{G}}^{\mu}} \phi(e_2)\\
          &\text{and}\\
          &\text{if } e_1 \textcolor{blue}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{blue}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \end{align*}
          \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$. 
  \end{enumerate}
  % If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
  \end{definition}
  The conditions on e-hypergraph homomorphisms require preserving immediate predecessors and the consistency relation.
  \begin{definition}[Category of e-hypergraphs]
  The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
  \end{definition}

The category of e-hypergraphs has coproducts,  given by the disjoint union of e-hypergraphs,  and an initial object given by the empty e-hypergraph.  
A concrete description of the pushout of two morphisms in this category is given in the Appendix section~\ref{sec:appendix:pushout}.
Generally,  the pushout of two e-hypergraph homomorphisms need not exist,  but it does when certain technical conditions are satisfied. 
Most importantly a pushout for the composition of two cospans with discrete feet exists, as described below.

As in the case of $\catname{S}(\Sigma)$ and $\MdaCospans$, e-hypergraphs requires interfaces in order to model string diagrams with dedicated inputs and outputs.
Because string diagrams for closed semilattice-enriched symmetric monoidal categories are nested, containing substring diagrams inside e-boxes and lambda abstraction boxes, we need to account for the interfaces in these nested contexts as well.
Nested inputs and outputs of a string diagram do not participate in composition of string diagrams, hence they need to be distinguished from the ones that participate in the composition.
We thus propose \textit{extended} cospans of discrete e-hypergraphs and use the notation $n \setminus m$ to denote the discrete e-hypergraph with $n \setminus m$ vertices, in particular
with the vertices of $m$ removed from $n$ when vertices of $m$ is a sub-e-hypergraph of $n$.

>>>>

First,  we fix some notation.  Let $(-)^*$ be the free monoid monad over \textsf{Set}.  
We extend this to the case of relations $R \subseteq V \times V$,  so that we denote by $R^{*} \subseteq V^* \times V^*$ the element-wise extension of $R$ to a relation on ordered sequences.
We will sometimes use functional notation for relations,  so that given $R \subseteq V \times W$ and $v \in V$,  $R(v) \subseteq W$. 
We elide associativity and unit isomorphisms associated with coproducts $(+)$,  and denote by $\iota_j: X_{j} \rightarrow X_{1} + \ldots X_{j} + \ldots X_{n}$ the $j^{\text{th}}$ injection,  and $[f,g]: A_1 + A_2 \to B$ the co-pairing of two morphisms $f: A_1 \to B$ and $g:A_2 \to B$.
We denote by $A +_{f,g} B$ the pushout of the span $A \xleftarrow{f} C \xrightarrow{g} B$.
We will also refer to $A,B$ as \textit{feet} of the cospan, and to $C$ as \textit{carrier}.
% Given an ordered sequence of labels from some alphabet $\Sigma$, $[l_1, \ldots, l_n]$ we will denote with $w([l_1, \ldots, l_n])$ the word concatenated from these labels using $\otimes$, \textit{e.g.}, $w([A,B]) = A \otimes B$.

@@

In ~\cite{bonchi_string_2022-2}, certain class of hypergraphs \textit{with interfaces} and their DPO rewriting are shown to be sound and complete for categorical presentations of arbitrary SMTs.
Hypergraphs generalise graphs by allowing edges to have multiple sources and targets. 
When interpreting string diagrams as hypergraphs,  morphisms are interpreted as edges,  and wires as vertices.
However,  there is also a need to specify which vertices are to be considered as \textit{input} and \textit{outputs},  corresponding to the dangling wires at the bottom and top of a string diagram. 
This gives rise to the definition of a hypergraph with interfaces.  
% Subsequent work \cite{bonchi_string_2022-2} builds on this correspondence to give a restriction of hypergraphs with interfaces (called \textit{monogamous} directed acyclic),  leading to a correspondence between symmetric monoidal string diagrams (\textit{i.e.}, without requiring a Frobenius algebra) and the category of (cospans of) such hypergraphs.
We then evoke the definition of hypergraphs and corresponding cospans as introduced in~\cite{bonchi_string_2022-1,bonchi_string_2022-2}.

\begin{definition}[Category of hypergraphs]\label{def:hypergraph}
A \emph{hypergraph $\mathcal{G}$ over monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$} is a tuple $(V,E,s,t,l)$,  where $V$ is a finite set of vertices, $E$ is a finite set of edges, $s : E \to V^{*}$ is a source function, $t : E \to V^{*}$ is a target function,  and $l = (l_{V} : V \to \Sigma_{O},l_{E} : E \to \Sigma_{M})$ is a pair of labelling functions that assigns each edge and vertex an operator and a type from monoidal signature $\Sigma$ respectively.  
The labelling function must respect the typing of the operator: for all edges $e$, such that $l(e) = c : A \to B$,  $l_{V}^{*}(s(e)) = A$ and $l_{V}^{*}(t(e)) = B$.  
We call a hypergraph \textit{discrete} if its set of edges is empty.   
A \emph{hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ is given by a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that the following hold. 
\begin{enumerate}
    \item $\phi_V^*(s_{\mathcal{F}}(e)) = s_{\mathcal{G}}(\phi_E(e)) \qquad \phi_V^*(t_{\mathcal{F}}(e)) = t_{\mathcal{G}}(\phi_E(e))$
    \item $l_{V,\mathcal{F}}(v) = l_{V,\mathcal{G}}(\phi_V(v)) \qquad\;\; l_{E,\mathcal{F}}(e) = l_{E,\mathcal{G}}(\phi_E(e))$
\end{enumerate}
That is, a homomorphism should preserve sources and targets of edges and preserve the labels.
We denote by $\catname{Hyp(\Sigma)}$ the category of hypergraphs over $\Sigma$ and hypergraph homomorphisms.
\end{definition}
$\catname{Hyp(\Sigma)}$ has all finite colimits and, in particular,  the coproduct is given by the disjoint union of hypergraphs.
The initial object is given by the empty hypergraph.

@@

% Then, morphisms of $\catname{S}(\Sigma)$ can be interpreted as particular cospans of discrete hypergraphs,  which we call hypergraphs with interfaces~\cite{bonchi_string_2022-2}.
\begin{definition}[Hypergraphs with interfaces]
\label{def:cspd}
A hypergraph with interfaces is a cospan $n \xrightarrow{f} \mathcal{G} \xleftarrow{g} m$ in $\catname{Hyp}(\Sigma)$ where $n$, $m$ are discrete ordered hypergraphs that denote interfaces.
Define the category of \emph{hypergraphs with interfaces} $\catname{HypI(\Sigma)}$ as the category with discrete \textit{ordered} hypergraphs as objects and isomorphism classes of hypergraphs with interfaces as morphisms.
\end{definition}
$\HypI{\Sigma}$ is a symmetric monoidal category with the tensor product given by coproduct in $\catname{Hyp}(\Sigma)$:
\begin{align*}
n \xrightarrow{f} &\mathcal{F} \xleftarrow{g} m\\
&\times \qquad \quad \Coloneqq n + n' \xrightarrow{f + f'} \mathcal{F} + \mathcal{G} \xleftarrow{g + g'} m + m'\\
n' \xrightarrow{f'} &\mathcal{G} \xleftarrow{g'} m'
\end{align*}
Unit is given by the initial object of $\catname{Hyp}(\Sigma)$.
Symmetry is inherited from coproduct in $\catname{Hyp}(\Sigma)$~\cite{MonoidalCoproduct}.

Composition of two cospans $n \xrightarrow{f} \mathcal{F} \xleftarrow{g} m;m \xrightarrow{f'} \mathcal{G} \xleftarrow{g'} k$ is given by the pushout
\[
n \xrightarrow{f} \mathcal{F} +_{g,f'} \mathcal{G} \xleftarrow{g'} k
\]
that is computed by identifying vertices in $\mathcal{F} + \mathcal{G}$ that share pre-image in $m$.
As pushout is only defined up to isomorphism, it justifies the morphisms in this category being isomorphism classes of cospans.

To make such cospans complete with respect to representing morphism of $\catname{S}(\Sigma)$, we need to impose some restriction on their form.
\begin{definition}[Degree,  path,  acyclic] 
The \emph{in-degree} of a vertex $v$ in a hypergraph $\mathcal{G}$ is the number of hyperedges $e \in {E_\mathcal{G}}$ such that $v \in t(e)$.  
Similarly, the out-degree of $v$ is the number of edges $e \in E_\mathcal{G}$ such that $v \in s(e)$.
A \emph{path} from $e_0$ to $e_{n-1}$ of length $n$ in a hypergraph $\mathcal{G}$ is an ordered sequence of edges $[e_0, e_1, \ldots, e_{n-1}] \in E_\mathcal{G}^*$ such that for all $i < n - 1$ some vertex $v \in t(e_i)$ is also in $s(e_{i+1})$.
A hypergraph is \emph{directed acyclic} if it contains no paths of length $n > 0$ that start and end in the same edge.
\end{definition}

@@

\begin{figure}[t!]
    \begin{subfigure}[T]{0.35\linewidth}
    \[
                % \tikzfig{../figures/combinatorial_semantics/DPO_square}
                \begin{tikzcd}
                    {\mathcal{L}} & {\mathcal{K}} & {\mathcal{R}} \\
                    {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}}
                    \arrow["f"', from=1-1, to=2-1]
                    \arrow[from=1-2, to=1-1]
                    \arrow[from=1-2, to=1-3]
                    \arrow[from=1-2, to=2-2]
                    \arrow[from=1-3, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
                    \arrow[from=2-2, to=2-1]
                    \arrow[from=2-2, to=2-3]
                    \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
                \end{tikzcd}
    \]
    % \subcaption{DPO square}
    % \label{fig:dpo}
    \end{subfigure}
    \hfill
    \begin{subfigure}[T]{0.4\linewidth}
        \[
    % \scalebox{0.8}{
        % \tikzfig{../figures/combinatorial_semantics/DPOI_square_Hyp}
    % }
    \begin{tikzcd}
        {\mathcal{L}} & {i+j} & {\mathcal{R}} \\
        {\mathcal{G}} & {\mathcal{L}^{\bot}} & {\mathcal{H}} \\
        & {n+m}
        \arrow["f"', from=1-1, to=2-1]
        \arrow[from=1-2, to=1-1]
        \arrow[from=1-2, to=1-3]
        \arrow[from=1-2, to=2-2]
        \arrow[from=1-3, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=90}, draw=none, from=2-1, to=1-2]
        \arrow[from=2-2, to=2-1]
        \arrow[from=2-2, to=2-3]
        \arrow["\lrcorner"{anchor=center, pos=0.125, rotate=180}, draw=none, from=2-3, to=1-2]
        \arrow[from=3-2, to=2-1]
        \arrow[from=3-2, to=2-2]
        \arrow[from=3-2, to=2-3]
    \end{tikzcd}
\]
    % \subcaption{DPOI square}
    % \label{fig:dpoi}
    \end{subfigure}
    \hfill
    \captionsetup{belowskip=-3ex}
    \caption{DPO and DPOI squares}
    \label{fig:dpo_dpoi}
\end{figure}

@@

Then, the authors of~\cite{bonchi_string_2022-2} showed that this enhanced notion of DPOI rewriting is sound and complete in the following sense.
Any two $\Sigma$-terms $f$ and $g$ are equal modulo SMC laws and equations $\mathcal{E}$ if and only if $\llbracket f \rrbracket \Rrightarrow{}_{\llbracket \mathcal{E} \rrbracket} \llbracket g \rrbracket$ where $\llbracket - \rrbracket$ is the functor giving the equivalence in Theorem~\ref{thm:prop-equiv} and $\llbracket \mathcal{E} \rrbracket = \{\langle \llbracket l \rrbracket, \llbracket r \rrbracket \rangle, \langle \llbracket r \rrbracket, \llbracket l \rrbracket \rangle \text{ for } l=r \in \mathcal{E} \}$.
We then proceed with building a combinatorial representation for closed $\Sigma^{+}$-terms to obtain a notion of DPOI rewriting that would also subsume equality modulo SMC laws.

\section{Closed E-hypergraphs}
Prior work of~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} introduced a notion of \textit{e-hypergraph} --- a combinatorial representation of morphism terms for free semilattice-enriched symmetric monoidal category with a modified notion of DPOI rewriting to make it sound and complete for equalities between morphism terms.
The semilattice enrichment structure was encoded using hierarchical edges in a hypergraph following the approach of a well-established area of hierarchical hypergraphs~\cite{plump:hierarchical-graphs, montanari:gs-lambda, palacz:hierarchical-transform, Gaducci:hierarchical-graphs, Ghica:hierarchical, fscd}.
The first work in \textit{loc. cit.}, in particular, introduced a combinatorial representation of closed $\Sigma$-terms (without enrichment).
Notably, the laws of a semilattice enrichment in~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} and the laws of Definition~\ref{def:closed} in~\cite{fscd} were not absorbed by the combinatorial representation and instead were implemented on top as structural DPOI rewrites.

@@

Consider an example in the middle part of Figure~\ref{fig:e-cospan-example}.
The e-hypergraph in the example is given by $V = \{v_{1}, \ldots w_{10}\}$, $E = \{e_{1}, \ldots, e_{7}\}$, $\textcolor{red}{E} = \{e_{1}\}$, $\textcolor{blue}{E} = \{e_{2}\}$, $l_{E} = {e_{1} \mapsto \bot, e_{2} \mapsto \bot, e_{3} \mapsto g, \ldots, e_{7} \mapsto f}$ and then we have
\[
\vspace{-1ex}
\begin{array}{cccc}
  v_{4} \textcolor{red}{<^{\mu}} e_{1} & v_{7} \textcolor{blue}{<^{\mu}} e_{2} & v_{4} \consistency e_{2} & v_{4} \not \consistency v_{10}\\
  e_{3} \textcolor{red}{<^{\mu}} e_{1} & v_{8} \textcolor{blue}{<^{\mu}} e_{2} &  e_{2} \consistency u_{1} & e_{2} \not \consistency e_{6}\\
  e_{2} \textcolor{red}{<^{\mu}} e_{1} & e_{5} \textcolor{blue}{<^{\mu}} e_{2} & e_{2} \consistency e_{4} & e_{4} \not \consistency e_{7}\\  
  \ldots & \ldots & \ldots & \ldots \\
  v_{12} \textcolor{red}{<^{\mu}} e_{1} & w_{9} \textcolor{blue}{<^{\mu}} e_{2} & v_{11} \consistency v_{12} & w_{6} \not \consistency w_{10}
\end{array}
\]
and, in particular, we have $[e_{5}) = \{e_2, e_1\}$.
$\consistency_{e_{1}}$ defines a partition on immediate successors of $e_{1}$ which we delimit by a vertical dashed line.
Edges in $\textcolor{red}{E}$ are depicted with a dashed box, while edges in $\textcolor{blue}{E}$ --- with a round solid box.
\begin{figure}
\[
\adjustbox{scale=0.6}{
\tikzfig{./figures/closed_iso_example_2}
}
\]
\captionsetup{skip=0pt, belowskip=-1ex}
\caption{Cospan of e-hypergraphs example}
\label{fig:e-cospan-example}
\end{figure}
\begin{definition}[E-hypergraph homomorphism]
  An \textit{e-hypergraph homomorphism} $\phi:\mathcal{F}\to\mathcal{G}$ between e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
  \begin{enumerate}
      \item $\phi$ is a hypergraph homomorphism;
      \item $\phi_{E}(\textcolor{red}{E_{\mathcal{F}}}) \subseteq \textcolor{red}{E_{\mathcal{G}}}$ and $\phi_{E}(\textcolor{blue}{E_{\mathcal{F}}}) \subseteq \textcolor{blue}{E_{\mathcal{G}}}$;
      \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
        \begin{align*}
          \text{if } e \textcolor{red}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{red}{<_{\mathcal{G}}^{\mu}} \phi(v)
          \qquad
          \text{if } e \textcolor{blue}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{blue}{<_{\mathcal{G}}^{\mu}} \phi(v)
        \end{align*}
          and for $e_1, e_2 \in E_{\mathcal{F}}$
          \begin{align*}
          \text{if } e_1 \textcolor{red}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{red}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \qquad
          \text{if } e_1 \textcolor{blue}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{blue}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \end{align*}
          \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$.
  \end{enumerate}
  % If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
  \end{definition}
  The conditions on e-hypergraph homomorphisms require preserving immediate predecessors and the consistency relation.
  \begin{definition}[Category of e-hypergraphs]
  The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
  \end{definition}

The category of e-hypergraphs has coproducts,  given by the disjoint union of e-hypergraphs,  and an initial object given by the empty e-hypergraph.  
A concrete description of the pushout of two morphisms in this category is given in the Appendix section~\ref{sec:appendix:pushout}.
Generally,  the pushout of two e-hypergraph homomorphisms need not exist,  but it does when certain technical conditions are satisfied. 
Most importantly a pushout for the composition of two cospans with discrete feet always exists, as described below.

As in the case of $\catname{S}(\Sigma)$ and $\MdaCospans$, e-hypergraphs require interfaces in order to model string diagrams with dedicated inputs and outputs.
Because string diagrams for closed semilattice-enriched symmetric monoidal categories are nested, containing substring diagrams inside e-boxes and lambda abstraction boxes, we need to account for the interfaces in these nested contexts as well.
Nested inputs and outputs of a string diagram do not participate in composition of string diagrams, hence they need to be distinguished from the ones that participate in the composition.
We thus propose \textit{extended} cospans of discrete e-hypergraphs and use the notation $n \setminus m$ to denote the discrete e-hypergraph with $n \setminus m$ vertices, in particular
with the vertices of $m$ removed from $n$ when vertices of $m$ is a sub-e-hypergraph of $n$.
